---
title: "analyzing-data-in-science-analysis"
format: html
editor: visual
knitr:
  opts_chunk:
    warning: false
    message: false
---

## Prepping, loading, setting up

```{r}
library(tidyverse)
library(googlesheets4)

d <- read_sheet("https://docs.google.com/spreadsheets/d/1gLyAAp6N_5g-Uo-eb61O-eW2apMzzqNdM8fLcBV7V3g/edit#gid=46458265",
                sheet = 1)

qual_q80 <- read_sheet("https://docs.google.com/spreadsheets/d/1gLyAAp6N_5g-Uo-eb61O-eW2apMzzqNdM8fLcBV7V3g/edit#gid=46458265",
                       sheet = "Q 80 - MASTER")

qual_q80 <- qual_q80 %>% janitor::clean_names()

names(qual_q80) <- str_c("qual_q80_", names(qual_q80))

names(qual_q80)[1] <- "response_id"

d <- d %>% 
  left_join(qual_q80)
```

## RQ 1

How do teachers support their students to analyze and interpret data?

### open coding - Q80 overall

qualitatively coded based on open/written responses to Q80

```{r}
overall_freqs <- d %>% 
  select(contains("qual_q80")) %>% 
  select(-c(1, 9)) %>% 
  summarize_all(sum, na.rm = T) %>% 
  gather(key, val) %>% 
  mutate(prop = val/328) %>% 
  arrange(desc(val))

overall_freqs %>% 
  mutate(prop = prop * 100) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  select(-val) %>% 
  knitr::kable()

overall_freqs %>% 
  ggplot(aes(x = reorder(key, prop), y = prop)) +
  geom_col() +
  theme_minimal() +
  coord_flip()
```

### fixed response question-- NSSME

4 overlapping
--
3 maybe overlapping/overlapping a little
2 not

Q98: How often do you have students do each of the following in your class?

1. Determine which data would need to be collected in order to answer a scientific question (regardless of who generated the question) (1) -- YES
2. Determine which variables from a provided dataset are necessary to answer a scientific question (13) -- MAYBE
3. Organize and/or represent data using tables, charts, or graphs in order to facilitate analysis (4) -- YES
4. Identify potential sources of variability (e.g., measurement error, natural variation) in the data (5) -- MAYBE
5. Analyze data using grade-appropriate methods in order to identify patterns, trends, or relationships (6) -- YES
6. Consider how missing data or measurement error can affect data interpretation (7) -- NOT
7. Select and use grade-appropriate mathematical and/or statistical techniques to analyze data (for example: determining the best measure of central tendency, examining variation in data, or developing a line of best fit) (9) -- YES
8. Use mathematical and/or computational models to generate data to support a scientific claim (10) -- NO
9. Use data and reasoning to defend, verbally or in writing, a claim or refute alternative scientific claims about a real-world phenomenon (regardless of who made the claims) (11) -- MAYBE

```{r}
d %>% 
  select(q98_1:q98_9) %>% 
  gather(key, val) %>% 
  count(key, val) %>% 
  filter(val != "NA") %>% 
  # Calculate total counts per key for percentage calculation
  group_by(key) %>% 
  mutate(total_n = sum(n)) %>% 
  ungroup() %>%
  # Calculate the percentage for each val
  mutate(percentage = (n / total_n) * 100) %>% 
  # Reorder based on "All or almost all science lessons"
  group_by(key) %>% 
  mutate(order_var = sum(n[val == "All or almost all science lessons"], na.rm = TRUE)) %>% 
  ungroup() %>%
  arrange(desc(order_var)) %>%
  # Use the label map for renaming keys
  mutate(key = fct_reorder(key, order_var, .desc = TRUE),
         val = factor(val, levels = c(
           "All or almost all science lessons", 
           "Often (once or twice a week)", 
           "Sometimes (once or twice a month)", 
           "Rarely (a few times a year)", 
           "Never"))) %>% 
  select(key, val, percentage) %>% 
  spread(val, percentage) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  knitr::kable()
```
## RQ2 - demographics

How does this support differe based on teachers’ grade, level of training and experience, and the subject they teach?

*we use what teachers say they are doing in their own words for RQ2 -- so we use Q80, not Q98*


### Grade Level

```{r}
d_grades <- d %>% 
  select(response_id, grade_level_taught) %>% 
  mutate(grade_level_taught = str_replace(grade_level_taught, ":", "")) %>% 
  separate(col = grade_level_taught, into = c(str_c("grade", letters[1:13])), sep = ",") %>% 
  gather(key, val, -response_id) %>% 
  arrange(response_id) %>% 
  mutate(val = as.integer(val)) %>% 
  mutate(elem = if_else(val <= 5, 1, 0)) %>% 
  mutate(second = if_else(val >= 6, 1, 0)) %>% 
  mutate(middle = if_else(val >= 6 & val <= 8, 1, 0)) %>% 
  mutate(high = if_else(val >= 9, 1, 0)) %>% 
  select(response_id, elem, second, middle, high) %>%
  group_by(response_id) %>% 
  summarize(elem = if_else(any(elem == 1), 1, 0),
            second = if_else(any(second == 1), 1, 0),
            middle = if_else(any(middle == 1), 1, 0),
            high = if_else(any(high == 1), 1, 0)) %>% 
  mutate_if(is.double, replace_na, 0)

d <- d %>% 
  left_join(d_grades)

d %>% 
  select(contains("qual_q80"), elem, middle, high) %>% 
  select(-c(1, 9)) %>% 
  mutate_all(replace_na, 0) %>%
  select(-middle, -high) %>% 
  group_by(elem) %>% 
  summarize_all(funs(sum), na.rm = TRUE) %>% 
  gather(key, val, -elem) %>% 
  filter(elem == 1) %>% 
  mutate(prop = val / 32) %>% 
  arrange(desc(prop)) %>% 
  knitr::kable()
```

```{r}
d %>% 
  select(contains("qual_q80"), elem, middle, high) %>% 
  select(-c(1, 9)) %>% 
  mutate_all(replace_na, 0) %>%
  select(-elem, -high) %>% 
  group_by(middle) %>% 
  summarize_all(funs(sum), na.rm = TRUE) %>% 
  gather(key, val, -middle) %>% 
  filter(middle == 1) %>% 
  mutate(prop = val / 114) %>% 
  mutate(prop = prop * 100) %>% 
  mutate(prop = round(prop, 2)) %>%
  arrange(desc(prop)) %>% 
  knitr::kable()
```

```{r}
d %>% 
  select(contains("qual_q80"), elem, middle, high) %>% 
  select(-c(1, 9)) %>% 
  mutate_all(replace_na, 0) %>%
  select(-elem, -middle) %>% 
  group_by(high) %>% 
  summarize_all(funs(sum), na.rm = TRUE) %>% 
  gather(key, val, -high) %>% 
  filter(high == 1) %>% 
  mutate(prop = val / 232) %>% 
  mutate(prop = prop * 100) %>% 
  mutate(prop = round(prop, 2)) %>%
  arrange(desc(prop)) %>% 
  knitr::kable()
```

### Subject area

```{r}
# Separate the values in the professional_role column into individual rows
d_long <- d %>%
  separate_rows(professional_role, sep = ",")

d_long <- d_long %>% 
  mutate(professional_role_6_text = stringr::str_to_lower(professional_role_6_text)) %>% 
  mutate(professional_role_6_text = str_detect(professional_role_6_text, "environmental")) # double-check this works
```

```{r}
# Trim any leading/trailing whitespace
d_long$professional_role <- trimws(d_long$professional_role)

# Create dichotomous columns for each unique professional role
d_wide <- d_long %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = professional_role, values_from = value, values_fill = 0)

d_wide <- d_wide %>% 
  select(contains("qual_q80"), Biology:Physics) %>% 
  select(-c(1, 9)) %>% 
  janitor::clean_names()

d_wide <- d_wide %>% 
  mutate_all(as.numeric) %>% 
  mutate_all(replace_na, 0)

# Identify task and subject columns
task_columns <- grep("^qual_q80", names(d_wide), value = TRUE)
subject_columns <- c("biology", "other_please_describe", "general_science",
                     "earth_science_or_earth_and_space_science", "chemistry", "physics")

# Function to calculate percentages for a single subject
calculate_percentages_by_subject <- function(subject_col) {
  total_subject <- sum(d_wide[[subject_col]] == 1, na.rm = TRUE)  # Total number of respondents for this subject
  
  d_wide %>%
    filter(!!sym(subject_col) == 1) %>%  # Only include rows where the subject was selected
    summarise(across(all_of(task_columns), ~ mean(. == 1, na.rm = TRUE) * 100)) %>%
    mutate(subject = subject_col, total_responses = total_subject)  # Add subject name and total responses
}

# Apply the function to all subject columns and bind the results
percentages_by_subject <- bind_rows(lapply(subject_columns, calculate_percentages_by_subject))

percentages_by_subject %>% 
  select(subject, total_responses, everything()) %>% 
  mutate_if(is.numeric, round, 2) %>%
  knitr::kable() # other is environmental
```

### years worked

**This tells us how these items correlate with years worked**

```{r}
d %>% 
  select(contains("qual_q80"), years_worked) %>% 
  select(-c(1, 9)) %>% 
  mutate(years_worked = unlist(years_worked)) %>% 
  mutate(years_worked = as.numeric(years_worked)) %>% 
  filter(years_worked < 100) %>% 
  mutate_if(is.numeric, replace_na, 0) %>% 
  corrr::correlate() %>% 
  corrr::focus(years_worked) %>% 
  arrange(desc(years_worked)) %>% 
  knitr::kable()
```

### prior experience

**CP: not sure this is helpful**

```{r}

d_long_prior <- d %>% 
  separate_rows(q98, sep = ",")

p <- d_long_prior %>% 
  select(contains("qual_q80"), prior_exp = q98) %>% 
  select(-c(1, 9)) %>% 
  mutate_if(is.numeric, replace_na, 0) %>% 
  gather(key, val, -prior_exp) %>%
  group_by(prior_exp, key) %>% 
  summarize(mean = mean(val)) %>% 
  mutate(mean = mean * 100)

p %>% 
  spread(key, mean) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  filter(prior_exp != "NA") %>% 
  knitr::kable()
```

### who does this - state (NGSS status)

```{r}
state_data <- read_sheet("https://docs.google.com/spreadsheets/d/1wul5jRNZBVqZP53swygLD34sBXPu0xEyJyOqRXLG6Dc/edit?gid=1207018323#gid=1207018323")

state_data <- state_data %>% 
  select(state = State, status = Status)

state_data

d %>% 
  select(contains("qual_q80"), state_work) %>%
  select(-c(1, 9)) %>% 
  rename(state = state_work) %>% 
  left_join(state_data, by = "state") %>% 
  mutate_if(is.numeric, replace_na, 0) %>% 
  gather(key, val, -status) %>% 
  group_by(status, key) %>% 
  mutate(val = as.integer(val)) %>% 
  summarize(mean = mean(val)) %>% 
  mutate(mean = mean * 100) %>% 
  spread(key, mean) %>% 
  filter(!is.na(status)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  select(-state)
  knitr::kable()
```
